---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  fig.width = 8.5,
  fig.height = 3,
  dpi = 250,
  message = FALSE
)
```
# daRt

<!-- badges: start -->
<!-- badges: end -->
This readme is very work in progress.

The daRt package provides a very quick and flexible way to import data that is produced by the Discrete Anisotropic Radiative Transfer (DART) model. The data in daRt are formatted in a way that facilitates rapid data analysis. 

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("willmorrison1/daRt")
```
## Example
Load the package
```{r load the package}
library(daRt)
```
Create and modify the "SimulationFilter" object. This defines what data you want to extract
from a DART output directory structure
```{r simulationFilter example}
#define SimulationFilter object - define "directions" as the product
sF <- simulationFilter(product = "directions")

#show the SimulationFilter
sF

#list the 'setters' and 'accessors'
methods(class = "SimulationFilter")

#e.g. change the 'bands', then the 'iterations'
bands(sF) <- c("BAND0", "BAND1")
iters(sF) <- "ITER1"
```
Now explore the DART output directory structure
```{r get files example}
#define the simulation directory
simulationDir <- "man/sampleSimulation/cesbio"

#define the SimulationFiler as shown above (i.e. 'sF'), but in one line
sF1 <- simulationFilter(product = "directions", 
                       bands = c("BAND0", "BAND1"), 
                       iters = "ITER1")
#get simulation files based on the defined filter
simFiles <- daRt::getFiles(x = simulationDir, sF = sF1)

#show these files are we happy to continue and load the data, or
#do we want to adjust the SimulationFilter? daRt::getFiles is essentially
#a 'dry-run' of the data extraction
files(simFiles)
```
Now extract DART output data
```{r get data example}
#get simulation data
simData <- daRt::getData(x = simulationDir, sF = sF1)
```
Documentation needs updating and finishing from here
```{r plot data example}
#plot using ggplot2
library(ggplot2)
plotOut <- ggplot(simData@data) +
    geom_point(aes(x = zenith, y = value, colour = azimuth)) +
    facet_wrap(~ band) +
    theme(aspect.ratio = 1)
plot(plotOut)
```
Then alter the SimulationFilter to look at images
```{r images example}
product(sF) <- "images"
simData <- daRt::getData(x = simulationDir, sF = sF)
ggplot(simData@data) + 
    geom_raster(aes(x = x, y = y, fill = value)) +
    facet_grid(band ~ imageNo) +
    theme(aspect.ratio = 1)
```
Alter the SimulatinFilter again to look at radiative budget files
```{r RB3D example}
product(sF) <- "rb3D"
simData <- daRt::getData(x = simulationDir, sF = sF)
ggplot(simData@data) + 
    geom_raster(aes(x = X, y = Y, fill = value)) +
    facet_grid(band + variablesRB3D~ Z) +
    theme(aspect.ratio = 1)
```
That's a lot of data! It is important to set the "SimulationFilter" to match what data you want as the process can use a lot of memory when many large files are loaded.
The below example uses the simple "dplyr" approach to work with the data. Here we look at the lowest horizontal layer of each 3D radiative budget array (i.e. Z = 1) and plot the smaller dataset.
```{r RB3D filter}
library(dplyr)
simData_filtered <- simData@data %>%
    dplyr::filter(Z == 1)
#plot again and tweak the plot
ggplot(simData_filtered) + 
    geom_raster(aes(x = X, y = Y, fill = value)) +
    facet_grid(band ~ variablesRB3D) +
    theme(aspect.ratio = 1) +
    theme_bw() +
    theme(panel.spacing = unit(0, "cm"), 
          strip.text = element_text(size = 6, 
                                    margin = margin(0.05, 0.05, 0.05, 0.05, unit = "cm"))) +
    theme(aspect.ratio = 1) +
    scale_fill_distiller(palette = "Spectral")
```
