---
output: 
  github_document:
    toc: true
    toc_depth: 2
title: daRt
---

<!-- README.md is generated from README.Rmd. Please edit and run README.Rmd file to regenerate README.md -->
```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%",
    fig.width = 8,
    fig.height = 3,
    dpi = 200,
    message = FALSE
)
```

<!-- badges: start -->
<!-- badges: end -->
The daRt package provides a very quick and flexible way to import data that is produced by the Discrete Anisotropic Radiative Transfer (DART) model. The data in daRt are formatted in a way that facilitates rapid data analysis. Formal documentation is available in the pdf manual.

# Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
# devtools::install_github doesn't like my github package dependencies 
remotes::install_github("willmorrison1/daRt")
```
Load the package
```{r load the package}
library(daRt)
```

# Overview
This section demonstrates the most basic use of daRt to load the "directions" data `product` for the default "cesbio" simulation provided in this respository.

Define a simulation directory
```{r}
simulationDir <- "man/data/cesbio"
```

**`simulationFilter()`** determines the type of files you want to load (here with defaults).
```{r}
sF <- daRt::simulationFilter(product = "directions")
```

**`getData()`** loads data for the given simulation using the predetermined file type(s).
```{r}
simData <- daRt::getData(x = simulationDir, sF = sF)
```

**`as.data.frame()`** releases the data object as a "long" format data frame.
```{r}
DF <- as.data.frame(simData)
head(DF, n = 3)
```

## Define data to load: SimulationFilter
The 'SimulationFilter' object describes what data you want to extract from a DART output directory structure.
Show the current configuration of the `SimulationFilter`
```{r}
sF
```
List the 'setter' and 'accessor' methods available
```{r}
methods(class = "SimulationFilter")
```
Use these methods to edit the `SimulationFilter` object e.g. the `bands` or `iters` (iterations) that you want to load
```{r}
bands(sF) <- 0:2
iters(sF) <- "ITER3"
```

## Browse files: SimulationFiles
The 'SimulationFiles' object contains all information on the files that will be loaded, based on the provided `SimulationFilter`. It is used to explore the DART output directory structure. 
First define the simulation directory. For this example, `simulationDir` is a relative directory (based on the github data provided) and consists of one simulation.
```{r}
#define the simulation directory
simulationDir <- "man/data/cesbio"
```
If you install the package using remotes::install_github then the "cesbio" simulation files will not be available automatically. To use these files, get them from github manually or use your own 'cesbio' simulation which is shipped with the DART model by default. 

The simulation directory should be the base directory of the simulation. E.g. within `simulationDir` there should be the simulation 'input' and 'output' directories.
```{r}
list.files(simulationDir)
```
Now we have the simulation directory clarified, explore the files in the simulation that correspond to this filter
```{r}
simFiles <- daRt::getFiles(x = simulationDir, sF = sF)
```
Explore the output of this to check we happy to continue and load the data. `getFiles()` is essentially a 'dry-run' of the data extraction
```{r}
dataFiles <- fileName(simFiles)
all(file.exists(dataFiles))
```

## Load data: SimulationData
The `SimulationData` object contains all data for the given `SimulationFilter`. Do the following to extract DART output data using the `getData()` method
```{r}
simData <- daRt::getData(x = simulationDir, sF = sF)
#also can do this using simFiles object
simData_fromFiles <- daRt::getData(x = simFiles)
identical(simData_fromFiles, simData)
```

## Analysis: as.data.frame()
By having data in a "long" format, it is easy to perform analysis on the data. Once you are ready to use the data, retrieve it using `as.data.frame()`.

```{r plot data example}
#plot using ggplot2
library(ggplot2)
DFdata <- as.data.frame(simData)
plotOut <- ggplot(DFdata) +
    geom_point(aes(x = zenith, y = value, colour = azimuth)) +
    facet_wrap(~ band) +
    theme(aspect.ratio = 1)
plot(plotOut)
```

# Expanded examples
This section provides further examples of package use.

## Defining data to load: SimulationFilter for images
To look at images for `bands` 0, 1 and 2; `iters` (iterations) 1 and 2, and `imageNums` (image numbers) 5 and 7, create the relevant SimulationFilter then load the data 
```{r images example}
#create SimulationFilter
sF <- daRt::simulationFilter(product = "images", 
                       bands = as.integer(0:2),
                       iters = c("ITER1", "ITER2"),
                       variables = "BRF",
                       imageNums = as.integer(c(5, 7)),
                       imageTypes = "ima") 
#load data - 'nCores' allows parallel processing of files.
#It is useful for access to drives that have optimised paralell I/O.
#here load data using 2 cores.
simData <- daRt::getData(x = simulationDir, sF = sF, nCores = 2)
#simple plot of data
ggplot(as.data.frame(simData)) + 
    geom_raster(aes(x = x, y = y, fill = value)) +
    facet_grid(band ~ imageNum + iter) +
    theme(aspect.ratio = 1) 
```

## Working with output data: radiative budget {#RB}
Alter the `SimulationFilter` again to now look at files for the radiative budget `product`.
```{r RB3D example}
product(sF) <- "rb3D"
simData <- daRt::getData(x = simulationDir, sF = sF, nCores = 2)
```

The 3D radiative budget data are stored with the X, Y and Z location of each cell (conforming to DART coordinate system i.e. "the part of the scene that horizontally is 'top left' and vertically is at the bottom is: X = 1, Y = 1, Z = 1), stored in 3 columns.
```{r RB3D data example}
head(as.data.frame(simData), n = 3)
```
The below example uses "dplyr" to work with this data. Here we look at the lowest horizontal layer of each 3D radiative budget array (i.e. Z = 1).
```{r}
library(dplyr)

#filter lowest horizontal cross section of the radiative budget
simData_filtered <- as.data.frame(simData) %>%
    dplyr::filter(Z == 1)

ggplot(simData_filtered) + 
    geom_raster(aes(x = X, y = Y, fill = value)) +
    facet_grid(band ~ variableRB3D) +
    theme_bw() +
    theme(panel.spacing = unit(0, "cm"), 
          strip.text = element_text(size = 6, 
                                    margin = margin(0.05, 0.05, 0.05, 0.05, unit = "cm"))) +
    scale_fill_distiller(palette = "Spectral") +
       theme(aspect.ratio = 1)
```

# Miscellaneous

## Memory management: tips
Loading many files/variables may require memory management. `getData()` loads all requested data to memory which can be problematic for large files (e.g. Radiative Budget). It is assumed that the user will perform some analysis on subsets of the raw data in a way that reduces the overall memory footprint. To demonstrate memory management, files in this section are loaded in two different scenarios: scenario 1 uses the default `getData()` to load and then analyse all data at once. Scenario 2 loads and analyses the data in pieces, which has a much smaller memory footprint (but may be slower). Both scenarios give the same result.

### Scenario 1: Load data all at once
Load all radiative budget products at once into memory and take the mean of each horizontal layer. 
```{r}
sF <- daRt::simulationFilter(product = "rb3D", 
                       bands = as.integer(0:2), 
                       iters = c("ITER1", "ITER2", "ILLUDIFF", "ILLUDIR"),
                       typeNums = "",
                       variables = "RADIATIVE_BUDGET")
simFiles <- daRt::getFiles(simulationDir, sF = sF)

```
There are twelve files each with 6 variables and each as a 3D array - i.e. quite a lot of data. Load in the data all at once.
```{r}
simData <- daRt::getData(x = simFiles, nCores = 2)
```
and gives a relatively large array of data
```{r}
DFdata <- as.data.frame(simData)
head(DFdata, n = 3)
dim(DFdata)
```
Do some analysis on the data. Get the mean of non-zero values across
each vertical layer of each `variablesRB3D`, `bands`, `iters` (already grouped) according to the above column names
```{r}
statVals <- DFdata %>%
    dplyr::group_by(X, Y, variableRB3D, add = TRUE) %>%
    dplyr::summarise(meanVal = mean(value[value != 0], na.rm = TRUE))
```

### Scenario 2: Load data in sections and process each section
Do 'scenario 1' analysis but with data processed for each band separately to save on memory usage.
```{r}
sF <- daRt::simulationFilter(product = "rb3D", 
                       bands = as.integer(0:2), 
                       iters = c("ITER1", "ITER2", "ILLUDIFF", "ILLUDIR"),
                       typeNums = "",
                       variables = "RADIATIVE_BUDGET")
allBands <- bands(simData)
allBands
simDataList <- vector(mode = "list", length = length(allBands))
for (i in 1:length(allBands)) {
    bands(sF) <- allBands[i]
    simDataPiece  <- daRt::getData(x = simulationDir, sF = sF, nCores = 2)
    simDataList[[i]] <- as.data.frame(simDataPiece) %>%
        dplyr::group_by(X, Y, variableRB3D, add = TRUE) %>%
        dplyr::summarise(meanVal = mean(value[value != 0], na.rm = TRUE))
}
```
Now put together the list of data. As each list element is a summary of the raw data, it has a much smaller memory footprint. As the summary was performed on one band at a time, the amount of data loaded at once is less than if `getData()` was executed for all bands at once (scenario 1). By loading one band at a time as opposed to all three at once, the memory footprint is around 1/3 of scenario 1. 
```{r}
simDataDF <- dplyr::bind_rows(simDataList)

statVals1 <- simDataDF

```
Both scenarios give the same results
```{r}
all.equal(statVals, statVals1)
```
but by processing in parts, the latter (scenario 2) - produced by 'statVals1' - has a smaller memory footprint as the stats are calculated for each band separately. When inter-band stats are required, the example can be adapted to iterate over e.g. `iters` or `variablesRB3D`.

## Radiative budget files: compression
DART radiative budget files are raw binary and can get very large. `rb3DtoNc` converts .bin to NetCDF (.nc) format, which gives smaller files sizes and can be compressed.

Get some DART radiative budget binary data (the default data)
```{r}
simulationDir <- "man/data/cesbio"
sF <- daRt::simulationFilter(product = "rb3D",
                       bands = as.integer(1), 
                       iters = "ITER1",
                       typeNums = "",
                       variables = "RADIATIVE_BUDGET")
simFiles_bin <- daRt::getFiles(simulationDir, sF = sF)
simData_bin <- as.data.frame(daRt::getData(simFiles_bin, nCores = 2))
#get the file size - for later comparison
fileSize_bin <- file.size(fileName(simFiles_bin))
```

```{r, echo = FALSE, results = 'hide',message = FALSE}
#crea te backup of the .bin file (it will be deleted by daRt::rb3DtoNc)
file.copy(fileName(simFiles_bin), "tmp.bin")
```
Convert the .bin data to .nc. The .bin file will be deleted by `rb3DtoNc`.
```{r}
simFiles_nc <- daRt::rb3DtoNc(simFiles_bin)
simData_nc <- as.data.frame(daRt::getData(simFiles_nc, nCores = 2))
```

There are some very minor differences in the two products - likely
due to the ncdf compression algorithm and/or rounding.

```{r}
max(abs(simData_nc$value - simData_bin$value))
```

The new .nc file is much smaller:
```{r}
fileSize_nc <- file.size(fileName(simFiles_nc))
fileSize_nc / fileSize_bin
```

and is much faster to read. It can also be read by third party NetCDF browsers e.g. ncview.

```{r, echo = FALSE, results = 'hide',message = FALSE}
#cleanup the creation of .nc
file.copy("tmp.bin", fileName(simFiles))
unlink("tmp.bin")
file.remove(fileName(simFiles_nc))
```

## Unwanted files: deletion
DART can output many unwanted files. Use `deleteFiles()` to delete files based on a provided `SimulationFiles` object. Here delete all `directions` files from `iters` 1 and `bands` 1. 

Define the file for deletion

```{r}
sF <- daRt::simulationFilter(product = "directions", 
                       bands = 1L, 
                       iters = "ITER1")
filesToDelete <- daRt::getFiles(x = simulationDir, sF = sF)
```

```{r, echo = FALSE, results = 'hide',message = FALSE}
#copy output files to save place
tmpFile <- paste0(fileName(filesToDelete), "_TMP")
file.copy(fileName(filesToDelete), tmpFile)
```

then delete the file, where `deleteSimulationFiles` parameter makes the user sure to 
know that they are deleting output data!

```{r}
deleteFiles(x = filesToDelete, deleteSimulationFiles = TRUE)

```


```{r, echo = FALSE, results = 'hide',message = FALSE}
#restore the deleted file
file.copy(tmpFile, gsub(pattern = "_TMP", replacement = "", tmpFile))
unlink(tmpFile)
```

## Parallelisation issues

Under windows especially, there may be some issues with paralellisation. When running `getData()` you may get "invalid connection" error. I may have sorted the issue but if it reappears then this function seems to help:
```{r}
unregister <- function() {
  env <- foreach:::.foreachGlobals
  rm(list = ls(name = env), pos = env)
}
```
